---
title: "Assignment week 10"
author: "Laurin Schaller"
date: "2023-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup and downloading resources

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT)
})
```


```{r, eval=FALSE}
options(timeout=999999)
download.file("https://ethz-ins.org/content/w10.assignment.zip", "w10.assignment.zip")
unzip("w10.assignment.zip")
```
# Importing peaks and regions

Unlike in the practical not all regions are in one file so we have to import them one by one

```{r}
tracks <- list.files(pattern="bw$")
peaks <- list.files(pattern="bed$")
peaks <- lapply(peaks, rtracklayer::import.bed)
# we'll focus on the high-quality peaks
peaks <- lapply(peaks, FUN=function(x) x[x$score>800])
# we get the union of non-redundant regions
regions <- reduce(unlist(GRangesList(peaks)))
```


```{r}

ml <- signal2Matrix(tracks, regions, extend=2000)
#nf <- getNormFactors(tracks, useSeqLevels="chr1")
#ml2 <- rescaleSignalMatrices(ml, scaleFactors=nf)
plotEnrichedHeatmaps(ml)
```
Can not get this to work, keep getting errors with range of k's:
Error in kmeans(dist(m), centers = k) :
'x' und 'centers' müssen dieselbe Zahl von Spalten haben

If I set a fixed k I get the following error:
Error in cl2$varExplained : $ operator is invalid for atomic vectors

```{r, eval=FALSE}
set.seed(123)
cl2 <- clusterSignalMatrices(ml, k=5)
ggplot(cl2$varExplained, aes(k, varExplained)) + geom_line()
```
