---
title: "Assignment Week 04"
author: "Laurin Schaller"
date: "2023-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading libraries

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(rtracklayer)
  library(epiwraps)
  library(GenomicRanges)
})
ah <- AnnotationHub()
```
# Data
All data chosen is from the same sets of experiments for consistency

- Mus musculus strain Bruce4 ES-Bruce4 cell line
- Bing Ren, UCSD

```{r, eval=FALSE}
options(timeout = max(999999, getOption("timeout")))
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM918750&format=file&file=GSM918750%5Fmm9%5FwgEncodeLicrTfbsEsb4P300ME0C57bl6StdPk%2EbroadPeak%2Egz", "encode/ep300.broadpeak.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM769008&format=file&file=GSM769008%5FwgEncodeLicrHistoneEsb4H3k4me3ME0C57bl6StdPk%2EbroadPeak%2Egz", "encode/H3K4me3.broadpeak.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM769009&format=file&file=GSM769009%5FwgEncodeLicrHistoneEsb4H3k4me1ME0C57bl6StdPk%2EbroadPeak%2Egz", "encode/H3K4me1.broadpeak.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM1000099&format=file&file=GSM1000099%5Fmm9%5FwgEncodeLicrHistoneEsb4H3k27acME0C57bl6StdPk%2EbroadPeak%2Egz", "encode/H3K27ac.broadpeak.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM1000089&format=file&file=GSM1000089%5Fmm9%5FwgEncodeLicrHistoneEsb4H3k27me3ME0C57bl6StdPk%2EbroadPeak%2Egz", "encode/H3K27me3.broadpeak.gz")
```

We have to use command line to unzip gz files
```{bash, eval=FALSE}
gunzip "encode/ep300.broadpeak.gz"
gunzip "encode/H3K4me3.broadpeak.gz"
gunzip "encode/H3K4me1.broadpeak.gz"
gunzip "encode/H3K27ac.broadpeak.gz"
gunzip "encode/H3K27me3.broadpeak.gz"
```
```{r}
list.files(path="encode/")
```

Files are looking good and are unzipped. Now we can load them

# Loading peaks
```{r}
peaksep300 <- rtracklayer::import("encode/ep300.broadpeak")
peaksH3K4me3<- rtracklayer::import("encode/H3K4me3.broadpeak")
peaksH3K4me1 <- rtracklayer::import("encode/H3K4me1.broadpeak")
peaksH3K27ac <- rtracklayer::import("encode/H3K27ac.broadpeak")
peaksH3K27me3 <- rtracklayer::import("encode/H3K27me3.broadpeak")
```

```{r}
head(peaksep300)
```
# Checking for overlaps
```{r}
p300OverlapsH3K4me3 <- overlapsAny(peaksep300, peaksH3K4me3)
p300OverlapsH3K4me1 <- overlapsAny(peaksep300, peaksH3K4me1)
p300OverlapsH3K27ac <- overlapsAny(peaksep300, peaksH3K27ac)
p300OverlapsH3K27me3 <- overlapsAny(peaksep300, peaksH3K27me3)
```

```{r}
head(p300OverlapsH3K4me3)
ol1 <- table(p300OverlapsH3K4me1)
ol1
ol2 <- table(p300OverlapsH3K4me3)
ol2
ol3 <- table(p300OverlapsH3K27ac)
ol3
ol4 <- table(p300OverlapsH3K27me3)
ol4
```
# Overlap data exploration

```{r}
totalOl <- sum(ol1[2], ol2[2], ol3[2], ol4[2])
totalOl
ol1[2]/totalOl * 100
ol2[2]/totalOl * 100
ol3[2]/totalOl * 100
ol4[2]/totalOl * 100
sumOverlaps <- c(ol1[2], ol2[2], ol3[2], ol4[2])
OlLabels <- c("H3K4me1", "H3K4me3", "H3K27ac", "H3K27me3")
pie(sumOverlaps, OlLabels, main="Overlaps of EP300 with Histone modifications", col = c("sienna1", "sienna2", "sienna3", "sienna4"))
```
Let's see where we have the most overlaps of EP300 with the four histone modifications. Also let's make it pretty.
As we can see from the pie chart close to half of the overlaps are with H3K4me1 and there are almost no overlaps with H3K27me3.

- The total overlaps are: 37364
- 49.38176% of overlaps are with H3K4me1
- 16.5159% of overlaps are with H3K4me3
- 32.91939% of overlaps are with H3K27ac
- 1.182957% of overlaps are with H3K27me3

# Biological background and interpretation
From the uniprot.org description of EP300:

- Functions as histone acetyltransferase and regulates transcription via chromatin remodeling. Acetylates all four core histones in nucleosomes.

From this the results in this analysis are a bit surprising to me. If it acetylates histones it makes sense that it interacts with methylated histones. But why is there such a big proportion of overlaps with H3K27ac?
Also the proportion of overlaps with H3K27me3 are rather small.
